---
title: "Land use classification using dtwSat"
author: "Victor Maus^[Institute for Geoinformatics, University of Münster, Heisenbergstraße 2, 48149 Münster, Germany], ^[National Institute for Space Research, Avenida dos Astronautas 1758, 12227010, São José dos Campos, Brazil.]"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
    bibliography: references.bib
---

<!-- 
# Build the README.md file for github webpage 
library(knitr)
knit(input="README.Rmd", output = "README.md")
-->

dtwSat
=====

### Time-Weighted Dynamic Time Warping for remote sensing time series analysis
dtwSat provides an implementation of Time-Weighted Dynamic Time Warping (TWDTW) method for satellite image time series analysis and land use classification. It is useful to account for natural and cultivated vegetation types even with inter-annual climatic and seasonal variability. Methods based on dynamic time warping are flexible to handle with irregular sampling and out of phase time series, and have achieved significant results in time series data mining. The [tutorial bellow][Land use classification using dtwSat] shows how to use the package for land use mapping using satellite image time series.

### Install
```{r installation, eval = FALSE}
devtools::install_github("vwmaus/dtwSat")
```

### Quick demo

This dome performs a TWDTW analysis and show the results.
```{r quick_demo, echo = TRUE, eval = TRUE, results = "hide", warning = FALSE, message = FALSE}
library(dtwSat, quietly = TRUE)
weight.fun = logisticWeight(alpha=-0.1, beta=100, theta=0.5)

alig = twdtw(x=waveletSmoothing(template.list[[1]]), 
             patterns=patterns.list, weight.fun = weight.fun, 
             span=180, keep=TRUE) 

class("twdtw")
print(alig)
summary(alig)
```

```{r plot-path, echo = TRUE, eval = TRUE, fig.path='figure/', fig.width=6, fig.height=8, fig.align='center', fig.cap='Typical temporal patterns.', warning = FALSE, message = FALSE}
library(ggplot2)
gp = plot(alig) + theme(text = element_text(size = 8, family="Helvetica"))
gp
```


Land use classification using dtwSat
=====
```{r child = 'vignettes/lucc.Rmd'}
```

