---
title: "1. Land use mapping using TWDTW-1NN and stars"
author: "Victor Maus"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{1. Land use mapping using TWDTW-1NN and stars}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: ./../inst/REFERENCES.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, dev = "png")
```

This vignette offers a concise guide for using version 1.0.0 or higher of the `dtwSat` package to generate a land-use map.
The package utilizes Time-Weighted Dynamic Time Warping (TWDTW) along with a 1-Nearest Neighbor (1-NN) classifier.
The subsequent sections will walk you through the process of creating a land-use map based on a set of training samples 
and a multi-band satellite image time series.

## Reading Training Samples

First, let's read a set of training samples that come with the `dtwSat` package installation.

```{r , echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
library(dtwSat)

samples <- st_read(system.file("mato_grosso_brazil/samples.gpkg", package = "dtwSat"), quiet = TRUE)
```

## Preparing the Satellite Image Time Series

The `dtwSat` package supports satellite images read into R using the `stars` package.
The installation comes with a set of MOD13Q1 images for a region within the Brazilian Amazon.
Note that timing is crucial for the TWDTW distance metric. To create a consistent image time series, 
we start by extracting the date of acquisition from the MODIS file names [@Didan:2015].

```{r , echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
tif_files <- dir(system.file("mato_grosso_brazil", package = "dtwSat"), pattern = "\\.tif$", full.names = TRUE)

acquisition_date <- as.Date(regmatches(tif_files, regexpr("[0-9]{8}", tif_files)), format = "%Y%m%d")

print(acquisition_date)
```

*Side note:* The date in the file name is not the true acquisition date for each pixel.
MOD13Q1 is a 16-day composite product, and the date in the file name is the first day of this 16-day period.

With the files and dates in hand, we can construct a stars satellite image time series for `dtwSat`.

```{r , echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
# read data-cube
dc <- read_stars(tif_files,
                 proxy = FALSE,
                 along = list(time = acquisition_date),
                 RasterIO = list(bands = 1:6))

# set band names
dc <- st_set_dimensions(dc, 3, c("EVI", "NDVI", "RED", "BLUE", "NIR", "MIR"))

# convert band dimension to attribute
dc <- split(dc, c("band"))

print(dc)
```

Note that it's important to set the date for each observation using the parameter `along`.
This will produce a 4D array (data-cube) that will be collapsed into a 3D array by converting 
the 'band' dimension into attributes. This prepares the data for training the TWDTW-1NN model.

## Create TWDTW-KNN1 model

```{r , echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
twdtw_model <- twdtw_knn1(x = dc,
                          y = samples,
                          cycle_length = 'year',
                          time_scale = 'day',
                          time_weight = c(steepness = 0.1, midpoint = 50),
                          formula = band ~ s(time))

print(twdtw_model)
```

In addition to the mandatory arguments `x` (satellite data-cube) and `y` (training samples), 
the TWDTW distance calculation also requires setting `cycle_length`, `time_scale`, and `time_weight`.
For more details, refer to the documentation using `?twdtw`. The argument `formula = band ~ s(time)` is optional.
If provided, training samples time sereis are resampled using Generalized Additive Models (GAMs), 
collapsing all samples with the same land-use label into a single sample. This reduces computational 
demands. The sample in the model can be visualized as follows:

```{r , echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
plot(twdtw_model)
```


## Land Use Prediction

Finally, we predict the land-use classes for each pixel location in the data-cube:

```{r , echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
lu_map <- predict(dc, model = twdtw_model)
print(lu_map)
```

The 'time' dimension was reduced to a single map. We can now visualize it using `ggplot`:

```{r , echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
ggplot() +
  geom_stars(data = lu_map) +
  theme_minimal()
```

Note that some pixels (in a 3x3 box) in the northeast part of the map have `NA` values due 
to a missing value in the blue band recorded on 2011-11-17. This limitation will be addressed 
in future versions of the dtwSat package.

Ultimately, we can write the map to a TIFF file we can use:

```{r , echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE}
write_stars(lu_map, "lu_map.tif")
```

## Further Reading

This introduction outlined the use of `dtwSat` for land-use mapping. 
For more in-depth information, refer to the papers by @Maus:2016 and @Maus:2019 and the 
[`twdtw` R package documentation](https://CRAN.R-project.org/package=twdtw).

For additional details on how to manage input and output satellite images, 
[`check` the stars documentation](https://CRAN.R-project.org/package=stars).

## References
